---
title: "Lipidomics pipeline - Required Code"
author: "Stefanie Warnat-Herresthal & Rainer Knoll"
date: "15/04/2020"
output: 
  html_document: 
    toc: true
    toc_float: true
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 5, message = F)
```


# 1. Load packages and functions
```{r, message=F}

library(SummarizedExperiment)
library(ggplot2)
library(tidyr)
library(pheatmap)
library(dplyr)
library(tsne)
library(pals)
library(DT)
library(RColorBrewer)
library(lubridate)
library(ggbeeswarm)
library(ggpubr)
library(readr)
library(plotly)
library(limma)
library(openxlsx)
library(reshape2)
library(ggrepel)

```

# 2. Functions
Load all functions necessary to run the script. If the file *lipidomics_functions.R* is located in the same directory as this script, it can be loaded like this. If not, specify the absolute path. 

```{r}
source("./lipidomics_functions.R")
```


# 3. Loading data and creating the lipids object

### Loading data
```{r}
lipidomics_data <- as.data.frame(read_csv("200110 COPD data new WF_AM adj to script_noPG-LPE-TAGodd.csv"))
sample_annotation <- read.csv("200110 Lipidomics COPD G2-4 anno.csv")
```

### Sample selection
By default, the script will all samples for downstream analysis, which are provided in the sample annotation. If you want to run your analysis only on specific samples, you can subset the object *sample_annotation* so that it only contains the samples of your interest. Only those will be taken into the analysis: 

```{r}
sample_annotation <- sample_annotation[!(sample_annotation$patient %in% c(59)),]
#sample exclusions due to pt59 control with very strong MAG signature (outlier)
```

Next, you can check whether the columns of your sample table are in the right format. This means that categorical variables (e.g. healthy vs. disease) should be factors, whereas numeric values should be numeric. Dates can be specified as dates or factors. 

```{r}
sample_annotation$sample <- as.factor(sample_annotation$sample)
sample_annotation$date_group<- as.factor(sample_annotation$date_group)
sample_annotation$group <- as.factor(sample_annotation$group)
sample_annotation$date <- lubridate::hms(sample_annotation$date)
#sample_annotation$stage <- as.factor(sample_annotation$stage)
sample_annotation$sex <- as.factor(sample_annotation$sex)
sample_annotation$age <- as.numeric(sample_annotation$age)
sample_annotation$BMI <- as.numeric(sample_annotation$BMI)
#sample_annotation$fixed <- as.factor(sample_annotation$fixed)
sample_annotation$smoking <- as.factor(sample_annotation$smoking)
sample_annotation$ex_smoking <- as.factor(sample_annotation$ex_smoking)
#sample_annotation$label <- as.factor(sample_annotation$label)
```

### Creating the lipid object
Now, we create the lipids object, which will store all data that we will calculate. It is created in every workflow before any data is subsetted or filtered. 

The function makeObject() creates two objects: 

1. *lipids*: *This is essentially the only one that you have to care about*. It contains all data from measured species. Inially, only the raw intensities are stored. They can be accessed via assay(lipids, "intensity"). Later, additional data (pmol, molpermil etc.) slots are added to the object. 

2. *standards*: contains all data from measured internal standards. Is only used to calculate the pmol values.

Both are of the class "SummarizedExperiment", which is a standard way of storing high-dimensional experimental data in R. For more background, see [Huber et al 2015](https://www.nature.com/articles/nmeth.3252): 


The arguments data and data_annotation take the intensity data and the sample annotation as described above. In addition, you can specify at which column of your Excel sheet the actual sample data starts, usually this is column 11.

```{r}
lip_obj <- makeObject(data = lipidomics_data, 
                      data_annotation = sample_annotation, 
                      sample_column_start = 11)
lipids <- lip_obj$lipids
standards <- lip_obj$standards
```

This function is used to convert all values to a tidy and easy plottable dataframe, the *summary*. This should be run after every operation on the lipids-object to update the summary. 

```{r}
summary <- getSummary()
```


### Colours

This chunk defines the colour scheme for plotting the annotations incl lipid classes. 
```{r}
# Sex
col_sex <- c("#b3151d", "#1643a6")
names(col_sex) <- c("F", "M")
# group
col_group <- c("#91bfdb", "#ffffbf", "#fc8d59")
names(col_group) <- c("ctrl", "2", "3_4")
# Smoking
col_smoking <- c("#a6d96a", "#f46d43", "#e3e3e3")
names(col_smoking) <- c("Y", "N", "NA")
# Ex smoking
col_ex_smoking <- c("#a6d96a", "#f46d43", "#e3e3e3")
names(col_ex_smoking) <- c("Y", "N", "NA")

# Date quartile
col_date_quartile <- c("#7b3294", "#c2a5cf", "#ffffbf", "#a6dba0", "#008837")
names(col_date_quartile) <- c("1_18", "2_18", "3_17", "3_18", "4_17")

palette_22 <- c("#771155", "#AA4488", "#CC99BB", "#114477", "#4477AA", "#77AADD", "#117777", "#44AAAA", "#77CCCC", "#117744", "#44AA77", "#88CCAA", "#777711", "#AAAA44", "#DDDD77", "#774411", "#AA7744", "#DDAA77", "#771122", "#AA4455", "#DD7788", "#606060")

n <- length(unique(rowData(lipids)$CLASS)) # number of colours

class_colours <- setNames(palette_22, unique(rowData(lipids)$CLASS))
anno_colours <- list(CLASS = class_colours,sex = col_sex, group = col_group, smoking=col_smoking, ex_smoking=col_ex_smoking, date_quartile = col_date_quartile) 
```

# 4. Intensity data 
The intensity data is the raw, unprocessed intensity data as provided in the Excel sheet. 

## QC - Overview tables {.tabset}
To examine the consistency of the dataset, you can check whether your data contains any NA or negative values.
### Undefined values
```{r}
print(paste("Number of undefined values in the dataset:", sum(is.na(summary$intensity))))
```
You can then eventually filter the species out that contain NA or negative  values. 

```{r}
keep <- complete.cases(assay(lipids, 1))
lipids <- lipids[keep,]
```

How many species remain in the dataset?
```{r}
dim(lipids)[1]
```

### Negative values
```{r}
print(paste("Number of negative values in the dataset:", sum(rowSums(assay(lipids, 1) < 0))))

print(paste("The following lipid species contain negative values and will be filtered out:", rownames(lipids)[rowSums(assay(lipids, 1) < 0)>1]))

filter <- rownames(lipids)[rowSums(assay(lipids, 1) < 0)>1]
lipids <- lipids[!rownames(lipids) %in% filter,]
```

Refresh summary
```{r}
summary <- getSummary()
```

# 5. Picomol data (absolute lipid content)

## Calculation
This chunk calculates the pmol from the intensity data by normalizing them based on the intensity of internal standards, which have known picomol values.

The calculated pmol values are added to the lipids-object. To access them, you can type assay(lipids, "pmol").

```{r, message=F}
lipids <- getPmol(lipid_object = lipids, filterNA = T)
summary <- getSummary() # adding pmol values to the summary object for plotting
```

# 6. Molpermil data (relative lipid content)

In this step, the pmol values are normalized on the total lipid content per sample, so that we can now compare relative differences across samples. 

By default, the data is normalized both on total lipid content as well as on a set of "stable" memnbrane lipids. Both results are added to the lipids object (slots "molpermil" and "molpermil_stable"). You can change the lipids that are conisered for the normalization on stable lipids via the argument "stable_classes" in the getMolpermil function. 

```{r}
lipids <- getMolpermil()
#filter <- rownames(lipids)[rowSums(assay(lipids, 1) < 0)>1]
#lipids <- lipids[!rownames(lipids) %in% filter,]
summary <- getSummary() # updated the summary object with the freshly calculated molpermil values
```


# -------------- Figure 5 B ---------------------


### Overall sum of molpermil values per condition

```{r, fig.width = 8, fig.height=4}
df <- summary %>% 
  group_by(CLASS, sample) %>% 
  summarize(class_wise_sum = sum(molpermil))
# add group annotation
merge_df <- summary[,c("sample", "group")]
merge_df <- distinct(merge_df)
df <- merge(df, merge_df, by = "sample")
df_mean <- df %>% 
  group_by(CLASS, group) %>% 
  summarize(mean = mean(class_wise_sum))

df_mean$group <- gsub(pattern = "2", replacement = "GOLD2", x = df_mean$group)
df_mean$group <- gsub(pattern = "3_4", replacement = "GOLD3/4", x = df_mean$group)
df_mean$group <- gsub(pattern = "ctrl", replacement = "Control", x = df_mean$group)

ggplot(df_mean, aes(x = group, y = mean)) + 
  geom_col(aes(fill = CLASS), width = 0.5, color ="black") + 
    scale_fill_manual(values =  as.character(anno_colours$CLASS)) + 
  guides(fill=guide_legend(ncol=2)) + 
  labs(y= "Normalized molpermil values", x = "")+
  theme_linedraw()+
  theme(panel.grid.major = element_line("white"),
        panel.grid.minor = element_line("white"),
        axis.title.y = element_text( size = 12),
        axis.text.y = element_text( size = 12),
        axis.title.x = element_text( size = 18),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
        legend.title = element_blank())+ 
  coord_flip()
  

```

# 7. Generating molpermil_sum table

```{r}
tmp <- as.data.frame(assay(lipids, "molpermil"))
tmp$NAME <- rownames(tmp)

tmp2 <- distinct(summary[,c("NAME", "CLASS")])

subset <- merge(tmp2, tmp, by = "NAME")

subset_melt <- melt(subset)
subset_melt <- subset_melt %>% 
  group_by(CLASS, variable) %>% 
  summarize(class_wise_sum = sum(value))

molpermil_sum <- as.data.frame(acast(subset_melt, CLASS~variable))

```


# -------------- Figure 5 C ---------------------


## Molpermil_sum per group

```{r, fig.width = 6, fig.height = 2}
names_2 <- as.character(sample_annotation[which(sample_annotation$group == "2"), c("group", "sample")]$sample)
names_3_4 <- as.character(sample_annotation[which(sample_annotation$group == "3_4"), c("group", "sample")]$sample)
names_ctrl <- as.character(sample_annotation[which(sample_annotation$group == "ctrl"), c("group", "sample")]$sample)

molpermil_sum_group <- molpermil_sum

molpermil_sum_group$"3_4" <- rowSums(molpermil_sum_group[,names_3_4])/length(names_3_4)
molpermil_sum_group$"2"  <- rowSums(molpermil_sum_group[,names_2])/length(names_2)
molpermil_sum_group$ctrl <- rowSums(molpermil_sum_group[,names_ctrl])/length(names_ctrl)

molpermil_sum_group <- molpermil_sum_group[, c("3_4", "2", "ctrl")]
molpermil_sum_group

heatmap_data <- molpermil_sum_group
colnames(heatmap_data) <- c("GOLD3/4", "GOLD2", "Control")
heatmap_data <- heatmap_data[,c(3,2,1)]

heatmap_data <- t(scale(t(heatmap_data)))
pheatmap(t(heatmap_data), 
         scale = "column", 
         cluster_rows = F,
         cluster_cols = T,
         show_rownames = T,
         show_colnames = T, 
         breaks = scaleColors(data = heatmap_data, maxvalue = 1.25)[["breaks"]], 
         color = colorRampPalette(c("#1b15c2", "#ffffff", "#ffbb00"))(length(scaleColors(data = heatmap_data, maxvalue = 1.25)[["breaks"]])), border_color = T, angle_col = 90)

```

# 8. Batch removal

Addition to the Lipidomics analysis to do a proper removal of the batch effect. 

## Step 1: Addition of log transformed data

```{r}
assay(lipids, 5) <- log2(as.data.frame(assay(lipids, 3))+0.01)
assayNames(lipids)[5] <- "molpermil_log"
```

## Step 2: Removal of batch effect

date_quartile was a batch effect so it will be removed.

```{r}
assay(lipids, 6) <- removeBatchEffect(as.data.frame(assay(lipids, 5)), 
                                      batch = droplevels(as.factor(colData(lipids)$date_quartile)), 
                                      design =model.matrix(~droplevels(as.factor(colData(lipids)$group))))
assayNames(lipids)[6] <- "molpermil_log_batchrm"

molpermil_log_batchrm <- assay(lipids, 6)
```

# 9. Differential Lipids

## Model matrix

Differential expression of lipids. First the model to be  fitted is sepecified. Since we compare the disease sate (Control, GOLD2 and GOLD3/4) the droplevels(colData(lipids)$group) is used. We observe a batch effect for date_quartile, therefore date_quartile is dadded in the design.  

```{r}
group_ <- droplevels(colData(lipids)$group)
date_ <- droplevels(colData(lipids)$date_quartile)
design <- model.matrix(~0 + group_ + date_)
```

## Contrast matrix

Define the comparisons of interest.

```{r}
contr.matrix <- makeContrasts(
   group_3_4_vs_group_ctrl = group_3_4 - group_ctrl, 
   group_2_vs_group_ctrl = group_2 - group_ctrl, 
   group_3_4_vs_group_2 = group_3_4 - group_2, 
   levels = colnames(design))
contr.matrix
```

## DL calling / lmFit

The log transformed molpermil data is used. lmFit fits a linear model using weighted least squares for each lipid. Estimate contrast for each lipid. And empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other lipids towards the average standard error) (see https://www.degruyter.com/doi/10.2202/1544-6115.1027)

```{r}
data <- assay(lipids, "molpermil_log")

vfit <- lmFit(data, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)

```

## Summary of DLs

Gives the number of up and down regulated lipids for all comparisons.

```{r}
dt <- decideTests(efit, adjust.method = "none")
summary(dt)
```

### GOLD3_4 vs Control

```{r}
top.table_GOLD3_4_vs_Control <- topTable(efit, sort.by = "P", n = Inf, adjust.method = "none", coef = "group_3_4_vs_group_ctrl")
top.table_GOLD3_4_vs_Control$lipid <- row.names(top.table_GOLD3_4_vs_Control)

top.table_GOLD3_4_vs_Control$Class <- "nope"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PC", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PCHO"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "CE", x = top.table_GOLD3_4_vs_Control$lipid)] <- "CHES"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "Cer", x = top.table_GOLD3_4_vs_Control$lipid)] <- "CERA"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "CL", x = top.table_GOLD3_4_vs_Control$lipid)] <- "CARD"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "DAG", x = top.table_GOLD3_4_vs_Control$lipid)] <- "DACG"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "HexCer", x = top.table_GOLD3_4_vs_Control$lipid)] <- "HEXC"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "DiHexCer", x = top.table_GOLD3_4_vs_Control$lipid)] <- "LACC"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "LPC", x = top.table_GOLD3_4_vs_Control$lipid)] <- "LPCH"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PC-O", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PC-O"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "MAG", x = top.table_GOLD3_4_vs_Control$lipid)] <- "MACG"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "LPC-O", x = top.table_GOLD3_4_vs_Control$lipid)] <- "LPCO"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PE", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PETA"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PE-O", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PE-O"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PI", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PHIN"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "PS", x = top.table_GOLD3_4_vs_Control$lipid)] <- "PHSE"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "SM", x = top.table_GOLD3_4_vs_Control$lipid)] <- "SPMY"
  top.table_GOLD3_4_vs_Control$Class[grepl(pattern = "TAG", x = top.table_GOLD3_4_vs_Control$lipid)] <- "TACG"


head(top.table_GOLD3_4_vs_Control, 20)
```

* logFC: log2 fold change of GOLD3/4 vs. Control 
* AveExpr: Average expression across all samples, here in log2 molpermil
* t: logFC divided by its standard error 
* P.Value: Raw p-value (based on t) from test that logFC differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
* B: log-odds that gene is DE (arguably less useful than the other columns)

### GOLD2 vs Control

```{r}
top.table_GOLD2_vs_Control <- topTable(efit, sort.by = "P", n = Inf, adjust.method = "none", coef = "group_2_vs_group_ctrl")
top.table_GOLD2_vs_Control$lipid <- row.names(top.table_GOLD2_vs_Control)

top.table_GOLD2_vs_Control$Class <- "nope"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PC", x = top.table_GOLD2_vs_Control$lipid)] <- "PCHO"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "CE", x = top.table_GOLD2_vs_Control$lipid)] <- "CHES"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "Cer", x = top.table_GOLD2_vs_Control$lipid)] <- "CERA"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "CL", x = top.table_GOLD2_vs_Control$lipid)] <- "CARD"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "DAG", x = top.table_GOLD2_vs_Control$lipid)] <- "DACG"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "HexCer", x = top.table_GOLD2_vs_Control$lipid)] <- "HEXC"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "DiHexCer", x = top.table_GOLD2_vs_Control$lipid)] <- "LACC"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "LPC", x = top.table_GOLD2_vs_Control$lipid)] <- "LPCH"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PC-O", x = top.table_GOLD2_vs_Control$lipid)] <- "PC-O"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "MAG", x = top.table_GOLD2_vs_Control$lipid)] <- "MACG"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "LPC-O", x = top.table_GOLD2_vs_Control$lipid)] <- "LPCO"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PE", x = top.table_GOLD2_vs_Control$lipid)] <- "PETA"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PE-O", x = top.table_GOLD2_vs_Control$lipid)] <- "PE-O"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PI", x = top.table_GOLD2_vs_Control$lipid)] <- "PHIN"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "PS", x = top.table_GOLD2_vs_Control$lipid)] <- "PHSE"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "SM", x = top.table_GOLD2_vs_Control$lipid)] <- "SPMY"
  top.table_GOLD2_vs_Control$Class[grepl(pattern = "TAG", x = top.table_GOLD2_vs_Control$lipid)] <- "TACG"

head(top.table_GOLD2_vs_Control, 20)
```


# Bubble plots

# -------------- Figure 5 C ---------------------

## COPD2 vs Control

```{r}
data <- top.table_GOLD2_vs_Control

ggplot(data, aes(x=lipid, y=logFC, size = P.Value)) +
   geom_point(aes(fill=Class), colour="black", pch=21) +
   scale_size("P.values", trans="log10", range=c(15, 4), breaks=c(0.015, 0.025, 0.5, 0.75, 1))+
   theme_bw() + 
   theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))+
   theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "none")+
   xlab("Species")+
  ggtitle("LogFC GOLD2 vs Control")+
  scale_fill_manual(values = as.character(anno_colours$CLASS)) 
```


# -------------- Figure 5 D ---------------------

## COPD3/4 vs Control

```{r, fig.width=6, fig.height=4}
data <- top.table_GOLD3_4_vs_Control

ggplot(data, aes(x=lipid, y=logFC, size = P.Value)) +
   geom_point(aes(fill=Class), colour="black", pch=21) +
   scale_size("P.values", trans="log10", range=c(15, 4), breaks= c(0.005, 0.01, 0.05,0.1, 0.5))+
   theme_bw() + 
   theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))+
   theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "none")+
   xlab("Species")+
  ggtitle("LogFC GOLD3/4 vs Control")+
  scale_fill_manual(values = as.character(anno_colours$CLASS)) 


```


# -----------------------------------------------------------------

# CoCena Part

# -----------------------------------------------------------------


# Preparations


```{r}
working_directory = paste0(setwd(dirname(rstudioapi::getSourceEditorContext()$path)),"/CoCena/")    

count_file_name = "molpermil_batchrm.csv"
topvar_genes = 2000

info_dataset_name = "sample_anno.csv"
voi = "group"

TF_list_name = "TFcat.txt"
gmtfile_name_hallmarks = "h.all.v6.1.symbols.gmt"
gmtfile_name_go = "c5.bp.v7.0.symbols.gmt"
gmtfile_name_kegg = "c2.cp.kegg.v7.0.symbols.gmt"
gmtfile_name_reactome = "c2.cp.reactome.v7.0.symbols.gmt"

organism = "human"

min_corr=0.5
range_cutoff_length=300
print_distribution_plots = FALSE

min_nodes_number_for_network=5
min_nodes_number_for_cluster=5

data_in_log=T

range_GFC=2.0

layout_algorithm = "layout_with_fr"  
```


# Packages

## Install packages:

```{r}
# install.packages(c("bench",
#                    "BiocManager",
#                    "combinat",
#                    "ComplexHeatmap",
#                    "dplyr",
#                    "ggnetwork",
#                    "ggplot2",
#                    "graphics",
#                    "grDevices",
#                    "grid",
#                    "gtools",
#                    "Hmisc",
#                    "igraph",
#                    "intergraph",
#                    "knitr",
#                    "MCDA",
#                    "openxlsx",
#                    "pals",
#                    "pheatmap",
#                    "purrr",
#                    "stringi",
#                    "tidyr",
#                    "tidyverse",
#                    "utils"))
# 
# 
# BiocManager::install(c("biomaRt",
#                        "clusterProfiler",
#                        "DOSE",
#                        "org.Hs.eg.db",
#                        "pcaGoPromoter.Hs.hg19",
#                        "RCy3",                       # Ensure that version 2.6.2. or newer is installed.
#                        "ReactomePA"))
```



## Load packages:

```{r}
library(clusterProfiler)
library(combinat)
library(ComplexHeatmap)
library(dplyr)
library(ggplot2) 
library(graphics)
library(grDevices)
library(grid)
library(igraph)
library(knitr)
library(pcaGoPromoter.Hs.hg19)
library(pheatmap)
library(purrr)
library(RColorBrewer)
library(stringi)
library(tidyr)
library(tidyverse)
library(utils)
library(ggnetwork)        
library(intergraph)      
library(MCDA)
```



# Data import

```{r}

# count_table <- read.table(paste0(working_directory, "data/", count_file_name) ,
#                          header = TRUE ,
#                          check.names = FALSE,
#                          sep = ",",
#                          row.names = 1,
#                          stringsAsFactors = F)


count_table <- read.table(paste0(working_directory, "data/", count_file_name) ,
                         header = TRUE ,
                         check.names = FALSE,
                         sep = ",",
                         row.names = 1,
                         stringsAsFactors = F)



info_dataset <- read.table(paste0(working_directory, "sample_info/", info_dataset_name),
                          header = TRUE ,
                          check.names = FALSE,
                          sep = ",",
                          row.names = 1,
                          stringsAsFactors = F)

rownames(info_dataset) <- info_dataset$sample

TF_list <- read.delim(paste0(working_directory, "reference_files/", TF_list_name),
                      header=TRUE,
                      check.names=F)

gmtfile_hallmarks <- clusterProfiler::read.gmt(paste0(working_directory, "reference_files/", gmtfile_name_hallmarks))
gmtfile_go <- clusterProfiler::read.gmt(paste0(working_directory, "reference_files/", gmtfile_name_go))
gmtfile_kegg <- clusterProfiler::read.gmt(paste0(working_directory, "reference_files/", gmtfile_name_kegg))
gmtfile_reactome <- clusterProfiler::read.gmt(paste0(working_directory, "reference_files/", gmtfile_name_reactome))
```



# Analysis

## Data filtering based on variability

Filtering data based on the intra-gene variance, will select first `r topvar_genes` genes based on their variance.

```{r}
ds = count_table[order(apply(count_table,1,var), decreasing=T),]
dd2 <- head(ds,topvar_genes)
dd2 = t(dd2)
```


## Correlation and correlation cut-off calculation

The first script *correlation_actions.R* calculates the **correlation_df** containing the information about the correlation of each gene to all other genes with respective p-value and r-value. Moreover, the script corrects for multiple testing, and filters the correlation_df for pval < 0.05 and r values > 0 resulting in **correlation_df_filt**. Additionally, the script calculates the cut-off values to be tested (**range_cutoff**).  

The second script *obtain_cutoff_stats.R* contains the function **cutoff_prep** for obtaining stats for each cutoff, namely R², the number of edges (no_edges), the number of nodes (no_nodes), as well as the number of networks (no_of_networks) the resulting igraph object would have.   
In addition, all degree distribution plots are saved in pdf format in a newly generated directory (if print_distribution_plots = T in *Preparations*). 

The third script *optimal_cutoff.R* calculates the optimal cut-off.

```{r}
source(paste0(working_directory,"scripts/", "correlation_actions.R"))

source(paste0(working_directory,"scripts/", "obtain_cutoff_stats.R"))

cutoff_stats = do.call("rbind", lapply(X = range_cutoff,
                                       FUN = cutoff_prep,
                                       corrdf_r = correlation_df_filt,
                                       print.all.plots = print_distribution_plots))

source(paste0(working_directory,"scripts/", "optimal_cutoff.R"))
```


Execute the following chunk to see the table containing all calculated cut-offs as well as to see the degree distribution plot for the calculated optimal cut-off.

```{r, results='asis'}
kable(cutoff_stats_concise, caption = "Correlation cut-off stats")
plot(dd_plot_calculated_optimal)
```


## Data filtering based on correlation cut off

**Set the optimal_cutoff your want to use for the further analysis.**  
A new directory for the chosen optimal_cutoff is generated in which all further results will be saved.   
The degree distribution plot for the chosen correlation cut-off is plotted and saved in the generated directory. In addition the count data is filtered (**filt_cutoff_counts**) to maintain only genes with at least one correlation to another gene above the chosen correlation cut-off. These genes are represented within a heatmap of the filtered count data.


```{r}

optimal_cutoff = calculated_optimal_cutoff

cutoff_wd <- paste0("dir_",optimal_cutoff, "_", topvar_genes)
if(!cutoff_wd %in% list.dirs(working_directory)) {
dir.create(paste0(working_directory,cutoff_wd))}

stats_optimal_cutoff <- cutoff_stats[cutoff_stats$cutoff == optimal_cutoff, c("degree", "Probs")]
dd_plot_optimal = ggplot(stats_optimal_cutoff,aes(x=log(degree), y= log(Probs))) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_bw() + 
  ggtitle(paste0("Chosen correlation cut-off [",optimal_cutoff, "]"))
print(dd_plot_optimal)
ggsave(filename = paste0("Degree_distribution_plot_", optimal_cutoff, ".pdf"), plot = dd_plot_optimal, 
       device = cairo_pdf, path = paste0(working_directory,cutoff_wd))


filt_cutoff_data = correlation_df_filt %>% dplyr::filter(rval > optimal_cutoff)
filt_cutoff_graph = igraph::graph_from_data_frame(filt_cutoff_data,directed=FALSE)
filt_cutoff_counts = ds[row.names(ds) %in% names(V(filt_cutoff_graph)),]
corresp_info = info_dataset[rownames(dd2)%in%rownames(info_dataset),]

print(paste("After using the optimal cutoff of",optimal_cutoff, "the number of edges =", 
            nrow(filt_cutoff_data), "and the number of nodes =", nrow(filt_cutoff_counts)))

heatmap_filtered_counts <- pheatmap::pheatmap(mat = filt_cutoff_counts,
                                              color=rev(RColorBrewer::brewer.pal(11, "RdBu")),
                                              scale="row",
                                              cluster_rows=T,
                                              cluster_cols=T,
                                              annotation_col=info_dataset[c(voi)],
                                              fontsize = 8,
                                              show_rownames = F, 
                                              show_colnames = F)

ggsave(filename = "Heatmap_topvar_genes.pdf", plot = heatmap_filtered_counts, device = cairo_pdf,
       path = paste0(working_directory,cutoff_wd))


```



## GFC calculation 

The group fold changes (GFCs) are calculated for each gene per condition. The defined variable of interest (voi) is used for the grouping of the samples.

```{r}

source(paste0(working_directory,"scripts/", "GFC_calculation.R" ))

GFC_all_genes <- GFC_calculation()


```


# -------------- Figure 5 F ---------------------


## Clustering

If *cluster_algo = "auto"*, five different cluster algorithms, namely cluster_label_prop, cluster_fast_greedy, cluster_louvain, cluster_infomap and cluster_walktrap are tested. Based on the highest modularity score, one of the algorithms is chosen for the actual clustering. Alternatively, one of the algorithms can be assigned to cluster_algo.  
In addition, the number of iterations can be specified (e.g. 100). If a gene is assigned to more clusters than specified in max_cluster_count_per_gene, it is painted white in the network, representing no definite cluster assignment. 

```{r, fig.width=6, fig.height=4}

source(paste0(working_directory,"scripts/", "cluster_calculation.R" ))

cluster_information <- cluster_calculation(igraph = filt_cutoff_graph,
                                           cluster_algo = "auto",
                                           no_of_iterations = 10,
                                           max_cluster_count_per_gene = 10,
                                           min_cluster_size = min_nodes_number_for_cluster)

source(paste0(working_directory,"scripts/", "heatmap_clusters.R" ))

cluster_information$color <- gsub(pattern = "darkgrey", replacement = "lightgreen", x = cluster_information$color)

heatmap_cluster <- heatmap_clusters(data = cluster_information)
 

```




## Network generation

Based on the filtered data, a network is calculated using the layout algorithm specified in the upper part of the script. In addition, all available node attributes are summarized in the data frame node_attributes.

```{r}

source(paste0(working_directory,"scripts/", "network_generation.R" ))

return_network <- network_layout(igraph.object = filt_cutoff_graph,
                                 use.layout = layout_algorithm,                        
                                 min.nodes.number = min_nodes_number_for_network)   

igraph_object <- return_network$graph_object
layout_matrix <- return_network$layout

node_attributes <- node_information(igraph.object = igraph_object,               
                                    data_df = cluster_information,
                                    GFC_df = GFC_all_genes,
                                    TF_df = TF_list,
                                    hallmark_df = gmtfile_hallmarks,
                                    go_df = gmtfile_go,
                                    kegg_df = gmtfile_kegg,
                                    reactome_df = gmtfile_reactome,
                                    org = organism)

network_object <- generate_network_object(graph_object = igraph_object,
                                          use_layout = layout_matrix)

```


# -------------- Figure E5 ----------------------


## Network visualization 

### Colored by Cluster

To visualize the **network colored by clusters**, execute the following chunk. The network will also be saved in pdf format, unless you set *save.pdf = F*.

```{r, fig.width=7, fig.height=5}
source(paste0(working_directory,"scripts/", "network_visualization.R" ))

network_cluster <- visualize_network(network = network_object,
                                     color.by = "cluster",
                                     select.cluster = NULL,
                                     plot.subnetwork = NULL,
                                     gene.label = c("PC(36:2)", 
                                                    "PC(38:5)", 
                                                    "CE(18:1)", 
                                                    "CE(20:4)",
                                                    "SM(22:1)", 
                                                    "LPC(22:6)",
                                                    "MAG(16:0)", 
                                                    "MAG(18:0)", 
                                                    "DAG(36:0)", 
                                                    "DAG(34:0)", 
                                                    "TAG(48:2)", 
                                                    "CE(22:1)",
                                                    "CE(22:1)", 
                                               "TAG(48:2)", 
                                               "MAG(16:0)", 
                                               "DAG(36:0)", 
                                               "MAG(18:0)", 
                                               "DAG(34:0)", 
                                               "CL(70:4)",
                                               "HexCer(d18:1/24:1)", 
                                               "HexCer(d18:1/16:0)", 
                                               "LPC-O(16:0)", 
                                               "PI(34:1)",
                                               "LPC(16:0)"),
                                     use.layout = layout_algorithm,
                                     save.pdf=T)
print(network_cluster)

```


# -------------- Figures 5 G and H ---------------------


### Lipids are highlighted

```{r, fig.width=7, fig.height=6}
network_cluster <- visualize_network(network = network_object,
                                     color.by = "cluster",
                                     select.cluster = c("orchid", "maroon"),
                                     plot.subnetwork = NULL,
                                     gene.label = c("PC(36:2)", 
                                                    "PC(38:5)", 
                                                    "CE(18:1)", 
                                                    "CE(20:4)",
                                                    "SM(22:1)", 
                                                    "LPC(22:6)"),
                                     use.layout = layout_algorithm,
                                     save.pdf=T)
print(network_cluster)


network_cluster <- visualize_network(network = network_object,
                                     color.by = "cluster",
                                     select.cluster = c("indianred", "lightgreen", "darkorange"),
                                     plot.subnetwork = NULL,
                                     gene.label = c("MAG(16:0)", 
                                                    "MAG(18:0)", 
                                                    "DAG(36:0)", 
                                                    "DAG(34:0)", 
                                                    "TAG(48:2)", 
                                                    "CE(22:1)","CL(70:4)"),
                                     use.layout = layout_algorithm,
                                     save.pdf=T)
print(network_cluster)

```




# Heatmap of selected Genes

Heatmap of genes from clusters/modules/selcted ones.

## List of Lipids from Clusters/Modules

```{r}
## getting genes out of cluster_information files
cluster_information <- as.data.frame(cluster_information)
subset <- cluster_information[,c("gene_n", "color")]
subset <- subset[subset$color != "white",]
subset <- as.data.frame(subset)
row.names(subset) <- subset$color
subset$color <- NULL
subset_genes <- separate(subset, gene_n, sep = ",", into= as.character(1:40))
row.names(subset_genes) <- row.names(subset)
subset_genes <- as.data.frame(t(subset_genes))

```

## Group means of all lipids

```{r}
filt_cutoff_counts_t <- as.data.frame(t(filt_cutoff_counts))
filt_cutoff_counts_t$group <- info_dataset[rownames(filt_cutoff_counts_t),"group"]
filt_cutoff_counts_t %>% group_by(group) %>% summarise_all(mean)  ->
  group_means 
group_means_df <- data.frame(t(group_means[,2:ncol(group_means)]),
                             row.names = colnames(group_means[,2:ncol(group_means)]))
colnames(group_means_df) <- group_means$group
group_means_df$Gene <- rownames(group_means_df)

head(group_means_df)
```


# -------------- Figures 5 I and J ---------------------


## heatmap clusters in for loop

```{r}
for (i in 1:length(colnames(subset_genes))){

groupmeans <- group_means_df[group_means_df$Gene %in% subset_genes[,i], ]
groupmeans$Gene <- NULL
groupmeans <- as.matrix(groupmeans)


breakList <- seq(-2, 2, by = .1)
pheatmap::pheatmap(mat = scale(t(groupmeans)),
                   scale = "column", 
                 #  breaks = breakList,
                   show_colnames = T,
                   labels_row = c("GOLD 2", "GOLD 3/4", "Control"),
                   cluster_rows = F, 
                   cluster_cols = T,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                 #  color = colorRampPalette(c("#1b15c2", "#ffffff", "#ffbb00"))(length(breakList)),
                 
         breaks = scaleColors(data = heatmap_data, maxvalue = 1.25)[["breaks"]], 
         color = colorRampPalette(c("#1b15c2", "#ffffff", "#ffbb00"))(length(scaleColors(data = heatmap_data, maxvalue = 1.25)[["breaks"]])),
                   border_color = T,
                   main = paste("Heatmap for module:", colnames(subset_genes)[i]))

print(pheatmap)

}

```

