---
title: "Fig.4D-F & Fig. E4B-C"
author: "Lisa Holsten"
date: "10 Juni 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages
```{r}
library(clusterProfiler)
library(dplyr)
library(ggplot2)
library(ggnetwork)
```

# Import data
```{r}
# Import info dataset
info_dataset <- read.table("sample_info/info_Dataset_Wataru_mod.csv", header = TRUE, check.names = FALSE, 
                           sep = ",", row.names = 1,stringsAsFactors = F)

# Import count table filtered for chosen cut-off (part of CoCena script)
filt_cutoff_counts <- readRDS(file = "CoCena_files/FilteredCountFile.rds")

# Import network (part of CoCena)
network_object <- readRDS(file = "CoCena_files/CoCena_NetworkObject.Rds")

# Import node attributes of CoCena network (generated within CoCena script)
node_attributes <- readRDS(file = "CoCena_files/NodeAttributes_CoCenaNetwork.rds")

# Import hallmark file
gmt_hallmark_entrez <- clusterProfiler::read.gmt("reference_files/h.all.v6.1.symbols.gmt")

# Import function for network visualization (part of CoCena)
source(paste0("scripts/network_visualization.R" ))
```

# Calculation of group means
```{r}
filt_cutoff_counts_t <- as.data.frame(t(filt_cutoff_counts))
filt_cutoff_counts_t$group <- info_dataset[rownames(filt_cutoff_counts_t),"condition"]
filt_cutoff_counts_t %>% group_by(group) %>% summarise_all(mean) -> group_means 

group_means_df <- data.frame(t(group_means[,2:ncol(group_means)]),
                             row.names = colnames(group_means[,2:ncol(group_means)]))
colnames(group_means_df) <- group_means$group
group_means_df$Gene <- rownames(group_means_df)
```

# List of genes of relevant clusters (> 35 and != white) & all network genes (universe):
```{r}
gene_list <- lapply(unique(node_attributes$cluster_color), function(x){
  unique(node_attributes[node_attributes$cluster_color==x,"Gene"])
})
names(gene_list) <- unique(node_attributes$cluster_color)

gene_list <- gene_list[lapply(gene_list,length)>=35]
gene_list$white = NULL

gene_list <-lapply(gene_list, function(x){
  gene_conv <- clusterProfiler:: bitr(x, 
                         fromType="SYMBOL", 
                         toType="ENTREZID", 
                         OrgDb="org.Hs.eg.db", 
                         drop = T)
  gene_conv$ENTREZID
})

universe_Entrez_network = clusterProfiler:: bitr(node_attributes$Gene, 
                                                 fromType="SYMBOL", 
                                                 toType="ENTREZID", 
                                                 OrgDb="org.Hs.eg.db", 
                                                 drop = T)
```

### compareCluster hallmark
```{r}
gmt_hallmark_entrez$ont <- stringr::str_replace(gmt_hallmark_entrez$ont, pattern="HALLMARK_", replacement = "")
gmt_hallmark_entrez$ont <- stringr::str_replace_all(gmt_hallmark_entrez$ont, pattern="_", replacement = " ")


colnames(gmt_hallmark_entrez) <- c("ont", "SYMBOL")

hallmark_entrez <- clusterProfiler:: bitr(gmt_hallmark_entrez$SYMBOL,
                                          fromType="SYMBOL",
                                          toType="ENTREZID",
                                          OrgDb="org.Hs.eg.db",
                                          drop = T)

gmt_hallmark_entrez <- dplyr::right_join(x = gmt_hallmark_entrez, y = hallmark_entrez, by = "SYMBOL")
gmt_hallmark_entrez <- gmt_hallmark_entrez[, -2]

ck_HALLMARK <- compareCluster(geneClusters = gene_list, fun = "enricher", pvalueCutoff = 1, qvalueCutoff = 1,
                              TERM2GENE=gmt_hallmark_entrez, universe=universe_Entrez_network$ENTREZID)

enrichplot::dotplot(ck_HALLMARK, showCategory = 5)
```


### Fig. 4D - Hallmark Cholesterol Homeostasis in GOLD2 modules
```{r}
ck_HALLMARK_df <- as.data.frame(ck_HALLMARK)

# Get gene names
genelist_cholesterol <- unlist(strsplit(x = ck_HALLMARK_df[ck_HALLMARK_df$Description == "CHOLESTEROL HOMEOSTASIS" & 
                                                             ck_HALLMARK_df$Cluster %in% c("gold", "red", "maroon"), 
                                                           "geneID"], 
                                 split = "/"))
genelist_cholesterol_symbol <- clusterProfiler:: bitr(genelist_cholesterol, 
                                               fromType="ENTREZID", 
                                               toType="SYMBOL", 
                                               OrgDb="org.Hs.eg.db", 
                                               drop = T)
genelist_cholesterol_symbol <- genelist_cholesterol_symbol$SYMBOL

# Network
visualize_network(network = network_object,
                  color.by = "cluster", 
                  select.cluster = c("gold", "red", "maroon"),
                  gene.label = genelist_cholesterol_symbol,
                  save.pdf = F) 

# Heatmap
groupmeans_HM_Chol <- group_means_df[group_means_df$Gene %in% genelist_cholesterol_symbol, ]
groupmeans_HM_Chol$Gene <- NULL
groupmeans_HM_Chol <- as.matrix(groupmeans_HM_Chol)

breakList <- seq(-1.5, 1.5, by = .1)
pheatmap::pheatmap(mat = groupmeans_HM_Chol,
                   scale = "row", 
                   breaks = breakList,
                   show_colnames = T,
                   labels_col = c("Control", "GOLD 2", "GOLD 3/4"),
                   cluster_rows = T, 
                   cluster_cols = F,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(length(breakList)-1))

```

### Fig. 4E - Hallmark Interferon Alpha Response in GOLD3/4 modules
```{r, fig.height=10}
# Get gene names
genelist_interferon_alpha <- unlist(strsplit(x = ck_HALLMARK_df[ck_HALLMARK_df$Description == "INTERFERON ALPHA RESPONSE" & 
                                                             ck_HALLMARK_df$Cluster %in% c("lightgreen", "pink"), 
                                                           "geneID"], 
                                        split = "/"))
genelist_interferon__alpha_symbol <- clusterProfiler:: bitr(genelist_interferon_alpha, 
                                                      fromType="ENTREZID", 
                                                      toType="SYMBOL", 
                                                      OrgDb="org.Hs.eg.db", 
                                                      drop = T)
genelist_interferon_alpha_symbol <- genelist_interferon__alpha_symbol$SYMBOL

# Network
visualize_network(network = network_object,
                  color.by = "cluster", 
                  select.cluster = c("lightgreen", "pink"),
                  gene.label = genelist_interferon_alpha_symbol,
                  save.pdf = F) 

# Heatmap
groupmeans_HM_INFA <- group_means_df[group_means_df$Gene %in% genelist_interferon_alpha_symbol, ]
groupmeans_HM_INFA$Gene <- NULL
groupmeans_HM_INFA <- as.matrix(groupmeans_HM_INFA)

breakList <- seq(-1.5, 1.5, by = .1)
pheatmap::pheatmap(mat = groupmeans_HM_INFA,
                   scale = "row", 
                   breaks = breakList,
                   show_colnames = T,
                   labels_col = c("Control", "GOLD 2", "GOLD 3/4"),
                   cluster_rows = T, 
                   cluster_cols = F,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(length(breakList)-1))

```


### Fig. 4F - Hallmark G2M checkpoint in GOLD3/4 modules
```{r, fig.height=8}
# Get gene names
genelist_G2M <- unlist(strsplit(x = ck_HALLMARK_df[ck_HALLMARK_df$Description == "G2M CHECKPOINT" & 
                                                            ck_HALLMARK_df$Cluster %in% c("lightgreen", "pink"), 
                                                          "geneID"], 
                                       split = "/"))
genelist_G2M_symbol <- clusterProfiler:: bitr(genelist_G2M, 
                                                     fromType="ENTREZID", 
                                                     toType="SYMBOL", 
                                                     OrgDb="org.Hs.eg.db", 
                                                     drop = T)
genelist_G2M_symbol <- genelist_G2M_symbol$SYMBOL

# Heatmap
groupmeans_HM_G2M <- group_means_df[group_means_df$Gene %in% genelist_G2M_symbol, ]
groupmeans_HM_G2M$Gene <- NULL
groupmeans_HM_G2M <- as.matrix(groupmeans_HM_G2M)

breakList <- seq(-1.5, 1.5, by = .1)
pheatmap::pheatmap(mat = groupmeans_HM_G2M,
                   scale = "row", 
                   breaks = breakList,
                   show_colnames = T,
                   labels_col = c("Control", "GOLD 2", "GOLD 3/4"),
                   cluster_rows = T, 
                   cluster_cols = F,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(length(breakList)-1))
```

### Fig. E4B - HAllmark Interferon gamma response
```{r, fig.height=11}
# Get gene names
genelist_IFNG <- unlist(strsplit(x = ck_HALLMARK_df[ck_HALLMARK_df$Description == "INTERFERON GAMMA RESPONSE" & 
                                                            ck_HALLMARK_df$Cluster %in% c("lightgreen", "pink"), 
                                                          "geneID"], 
                                       split = "/"))
genelist_IFNG_symbol <- clusterProfiler:: bitr(genelist_IFNG, 
                                                     fromType="ENTREZID", 
                                                     toType="SYMBOL", 
                                                     OrgDb="org.Hs.eg.db", 
                                                     drop = T)
genelist_IFNG_symbol <- genelist_IFNG_symbol$SYMBOL

# Heatmap
groupmeans_HM_IFNG <- group_means_df[group_means_df$Gene %in% genelist_IFNG_symbol, ]
groupmeans_HM_IFNG$Gene <- NULL
groupmeans_HM_IFNG <- as.matrix(groupmeans_HM_IFNG)

breakList <- seq(-1.5, 1.5, by = .1)
pheatmap::pheatmap(mat = groupmeans_HM_IFNG,
                   scale = "row", 
                   breaks = breakList,
                   show_colnames = T,
                   labels_col = c("Control", "GOLD 2", "GOLD 3/4"),
                   cluster_rows = T, 
                   cluster_cols = F,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(length(breakList)-1))
```


### Fig. E4C - Hallmark Apical Junctions in control modules
```{r, fig.height=10}
#Get genes names
genelist_apicaljunctions <- unlist(strsplit(x = ck_HALLMARK_df[ck_HALLMARK_df$Description == "APICAL JUNCTION" & 
                                                                 ck_HALLMARK_df$Cluster %in% c("orchid", "blue", 
                                                                                               "lightblue", "darkgrey",
                                                                                               "darkorange", "darkgreen"), 
                                                               "geneID"], split = "/"))
genelist_apicaljunctions_symbol <- clusterProfiler:: bitr(genelist_apicaljunctions, 
                                                     fromType="ENTREZID", 
                                                     toType="SYMBOL", 
                                                     OrgDb="org.Hs.eg.db", 
                                                     drop = T)
genelist_apicaljunctions_symbol <- genelist_apicaljunctions_symbol$SYMBOL

# Network
visualize_network(network = network_object,
                  color.by = "cluster", 
                  select.cluster = c("orchid", "blue", "lightblue", "darkgrey", "darkorange", "darkgreen"),
                  gene.label = genelist_apicaljunctions_symbol,
                  save.pdf = F) 

# Heatmap
groupmeans_HM_AJ <- group_means_df[group_means_df$Gene %in% genelist_apicaljunctions_symbol, ]
groupmeans_HM_AJ$Gene <- NULL
groupmeans_HM_AJ <- as.matrix(groupmeans_HM_AJ)

breakList <- seq(-1.5, 1.5, by = .1)
pheatmap::pheatmap(mat = groupmeans_HM_AJ,
                   scale = "row", 
                   breaks = breakList,
                   show_colnames = T,
                   labels_col = c("Control", "GOLD 2", "GOLD 3/4"),
                   cluster_rows = T, 
                   cluster_cols = F,
                   clustering_distance_rows = "euclidean",
                   clustering_method = "complete",
                   treeheight_row = 10,
                   angle_col = 90,
                   cellheight = 12,
                   cellwidth = 12,
                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(length(breakList)-1))
```
